```{r msscov-setup, include=FALSE, purl=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "msscov-")
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache=TRUE, tidy.opts=list(width.cutoff=60), tidy=TRUE, fig.align='center', out.width='80%')
```

# MARSS models with covariates {#chap-msscov}
\chaptermark{MARSS with covariates}

A script with all the R code in the chapter can be downloaded  [here](./Rcode/multivariate-ss-cov.R). The Rmd for this chapter can be downloaded [here](./Rmds/multivariate-ss-cov.Rmd)

### Data and packages {-}

For the chapter examples, we will use the green and bluegreen algae in the Lake Washington plankton data set and the covariates in that dataset.  This is a 32-year time series (1962-1994) of monthly plankton counts (cells per mL) from Lake Washington, Washington, USA with the covariates total phosphorous and pH.  `lakeWAplanktonTrans` is a transformed version of the raw data used for teaching purposes. Zeros have been replaced with NAs (missing). The logged (natural log) raw plankton counts have been standardized to a mean of zero and variance of 1 (so logged and then z-scored). Temperature, TP and pH were also z-scored but not logged (so z-score of the untransformed values for these covariates). The single missing temperature value was replaced with -1 and the single missing TP value was replaced with -0.3.

We will use the 10 years of data from 1965-1974 (Figure \@ref(fig:msscov-plank-plot)), a decade with particularly high green and bluegreen algae levels. 
```{r msscov-load-plankton-data}
data(lakeWAplankton, package="MARSS")
# lakeWA
fulldat = lakeWAplanktonTrans
years = fulldat[,"Year"]>=1965 & fulldat[,"Year"]<1975
dat = t(fulldat[years,c("Greens", "Bluegreens")])
covariates = t(fulldat[years,c("Temp", "TP")])
```

Packages:
```{r msscov-packages}
library(MARSS)
library(ggplot2)
```

## Overview {#sec-msscov-overview}

A multivariate autoregressive state-space (MARSS) model with covariate effects in both the process and observation components is written as:
\begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}_t\mathbf{x}_{t-1} + \mathbf{u}_t + \mathbf{C}_t\mathbf{c}_t + \mathbf{w}_t, \text{ where } \mathbf{w}_t \sim \text{MVN}(0,\QQ_t) \\
\mathbf{y}_t = \ZZ_t\mathbf{x}_t + \mathbf{a}_t + \mathbf{D}_t\mathbf{d}_t + \mathbf{v}_t, \text{ where } \mathbf{v}_t \sim \text{MVN}(0,\mathbf{R}_t) 
\end{gathered}
(\#eq:msscov-covars)
\end{equation}
where $\mathbf{c}_t$ is the $p \times 1$ vector of covariates (e.g., temperature, rainfall) which affect the states and $\mathbf{d}_t$ is a $q \times 1$ vector of covariates (potentially the same as $\mathbf{c}_t$), which affect the observations.   $\mathbf{C}_t$ is an $m \times p$ matrix of coefficients relating the effects of $\mathbf{c}_t$ to the $m \times 1$ state vector $\mathbf{x}_t$, and $\mathbf{D}_t$ is an $n \times q$ matrix of coefficients relating the effects of $\mathbf{d}_t$ to the $n \times 1$ observation vector $\mathbf{y}_t$.

With the `MARSS()` function, one can fit this model by passing in `model$c` and/or `model$d` in the `model` argument as a $p \times T$ or $q \times T$ matrix, respectively.  The form for $\mathbf{C}_t$ and $\mathbf{D}_t$ is similarly specified by passing in `model$C` and/or `model$D`.   $\mathbf{C}$ and $\mathbf{D}$ are matrices and are specified as 2-dimensional matrices as you would other parameter matrices.

## Prepare the plankton data {#sec-msscov-prepare-data}

We will prepare the data by z-scoring. The original data `lakeWAplanktonTrans` were already z-scored, but  we changed the mean when we subsampled the years so we need to z-score again. 

```{r msscov-z-score-data}
# z-score the response variables
the.mean = apply(dat,1,mean,na.rm=TRUE)
the.sigma = sqrt(apply(dat,1,var,na.rm=TRUE))
dat = (dat-the.mean)*(1/the.sigma)
```
Next we set up the covariate data, temperature and total phosphorous. We z-score the covariates to standardize and remove the mean.
```{r msscov-z-score-covar-data}
the.mean = apply(covariates,1,mean,na.rm=TRUE)
the.sigma = sqrt(apply(covariates,1,var,na.rm=TRUE))
covariates = (covariates-the.mean)*(1/the.sigma)
```

(ref:msscov-plank-dat) Time series of Green and Bluegreen algae abundances in Lake Washington along with the temperature and total phosporous covariates.

```{r msscov-plank-plot, fig=TRUE, echo=FALSE, fig.cap='(ref:msscov-plank-dat)', warning=FALSE}
LWA <- ts(cbind(t(dat), t(covariates)), start=c(1965,1), end=c(1974,12), freq=12)
plot(LWA, main="", yax.flip=TRUE)
```


## Observation-error only model {#sec-msscov-obs-error-only}

We can estimate the effect of the covariates using a process-error only model, an observation-error only model, or a model with both types of error.  An observation-error only model is a multivariate regression, and we will start here so you see the relationship of MARSS model to more familiar linear regression models.

In a standard multivariate linear regression, we only have an observation model with independent errors (the state process does not appear in the model):
\begin{equation}
\mathbf{y}_t = \mathbf{a} + \mathbf{D}\mathbf{d}_t + \mathbf{v}_t, \text{ where } \mathbf{v}_t \sim \text{MVN}(0,\mathbf{R}) 
(\#eq:msscov-y)
\end{equation}
The elements in $\mathbf{a}$ are the intercepts and those in $\mathbf{D}$ are the slopes (effects).  We have dropped the $t$ subscript on $\mathbf{a}$ and $\mathbf{D}$ because these will be modeled as time-constant.  Writing this out for the two plankton and the two covariates we get:
\begin{equation}
\begin{split}
 \begin{bmatrix}
    y_{g} \\
    y_{bg} \end{bmatrix}_t &=
\begin{bmatrix}
    a_1 \\
    a_2 \end{bmatrix} +
 \begin{bmatrix}
    \beta_{\mathrm{g,temp}}&\beta_{\mathrm{g,tp}} \\
    \beta_{\mathrm{bg,temp}}&\beta_{\mathrm{bg,tp}}  \end{bmatrix} 
 \begin{bmatrix}
    \mathrm{temp} \\
    \mathrm{tp} \end{bmatrix}_{t-1} +
 \begin{bmatrix}
    v_{1} \\
    v_{2} \end{bmatrix}_t 
\end{split}   
(\#eq:msscov-y-mat)
\end{equation}

Let's fit this model with MARSS.  The $\mathbf{x}$ part of the model is irrelevant so we want to fix the parameters in that part of the model.  We won't set $\mathbf{B}=0$ or $\ZZ=0$ since that might cause numerical issues for the Kalman filter. Instead we fix them as identity matrices and fix $\mathbf{x}_0=0$ so that $\mathbf{x}_t=0$ for all $t$.

```{r msscov-covar-model-0 }
Q <- U <- x0 <- "zero"; B <- Z <- "identity"
d <- covariates
A <- "zero"
D <- "unconstrained"
y <- dat # to show relationship between dat & the equation
model.list <- list(B=B,U=U,Q=Q,Z=Z,A=A,D=D,d=d,x0=x0)
kem <- MARSS(y, model=model.list)
```

We set `A="zero"` because the data and covariates have been demeaned.   Of course, one can do multiple regression in R using, say, `lm()`, and that would be much, much faster.  The EM algorithm is over-kill here, but it is shown so that you see how a standard multivariate linear regression model is written as a MARSS model in matrix form.


## Process-error only model {#sec-msscov-proc-error-only}

Now let's model the data as an autoregressive process observed without error, and incorporate the  covariates into the process model.  Note that this is much different from typical linear regression models. The $\mathbf{x}$ part represents our model of the data (in this case plankton species).  How is this different from the autoregressive observation errors? Well, we are modeling our data as autoregressive so data at $t-1$ affects the data at $t$.  Population abundances are inherently autoregressive so this model is a bit closer to the underlying mechanism generating the data.  Here is our new process model for plankton abundance.
\begin{equation}
\mathbf{x}_t = \mathbf{x}_{t-1} +  \mathbf{C}\mathbf{c}_t + \mathbf{w}_t, \text{ where } \mathbf{w}_t \sim \text{MVN}(0,\QQ) 
(\#eq:msscov-x)
\end{equation}
We can fit this as follows:

```{r msscov-covar-model-1 }
R <- A <- U <- "zero"; B <- Z <- "identity"
Q <- "equalvarcov"
C <- "unconstrained"
model.list <- list(B=B,U=U,Q=Q,Z=Z,A=A,R=R,C=C,c=covariates)
kem <- MARSS(dat, model=model.list)
```

Now, it looks like temperature has a strong negative effect on algae?  Also our log-likelihood dropped a lot.  Well, the data do not look at all like a random walk model (where $\mathbf{B}=1$), which we can see from the plot of the data (Figure \@ref(fig:msscov-plank-plot)).  The data are fluctuating about some mean so let's switch to a better autoregressive model---a mean-reverting model.  To do this, we will allow the diagonal elements of $\mathbf{B}$ to be something other than 1.

```{r msscov-covar-model-1c }
model.list$B <- "diagonal and unequal"
kem <- MARSS(dat, model=model.list)
```

Notice that the log-likelihood goes up quite a bit, which means that the mean-reverting model fits the data much better.

With this model,  we are estimating $\mathbf{x}_0$.  If we set `model$tinitx=1`, we will get a error message that $\mathbf{R}$ diagonals are equal to 0 and we need to fix `x0`.  Because $\mathbf{R}=0$, if we set the initial states at $t=1$, then they are fully determined by the data.

```{r msscov-covar-model-2 }
x0 <- dat[,1,drop=FALSE]
model.list$tinitx <- 1
model.list$x0 <- x0
kem <- MARSS(dat, model=model.list)
```


## Both process- and observation-error {#sec-msscov-both-error}

Here is an example where we have both process and observation error but the covariates only affect the process:
\begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}\mathbf{x}_{t-1} +  \mathbf{C}_t\mathbf{c}_t + \mathbf{w}_t, \text{ where } \mathbf{w}_t \sim \text{MVN}(0,\QQ)\\ 
\mathbf{y}_t = \mathbf{x}_{t-1} +   \mathbf{v}_t, \text{ where } \mathbf{v}_t \sim \text{MVN}(0,\mathbf{R}), 
\end{gathered}
(\#eq:msscov-xy-process-only)
\end{equation}
$\mathbf{x}$ is the true algae abundances and $\mathbf{y}$ is the observation of the $\mathbf{x}$'s.

Let's say we knew that the observation variance on the algae measurements was about 0.16 and we wanted to include that known value in the model.  To do that, we can simply add $\mathbf{R}$ to the model list from the process-error only model in the last example.

```{r msscov-covar-model-5 }
D <- d <- A <- U <- "zero"; Z <- "identity"
B <- "diagonal and unequal"
Q <- "equalvarcov"
C <- "unconstrained"
c <- covariates
R <- diag(0.16,2)
x0 <- "unequal"
tinitx <- 1
model.list <- list(B=B,U=U,Q=Q,Z=Z,A=A,R=R,D=D,d=d,C=C,c=c,x0=x0,tinitx=tinitx)
kem <- MARSS(dat, model=model.list)
```

Note, our estimates of the effect of temperature and total phosphorous are not that different than what you get from a simple multiple regression (our first example).  This might be because the autoregressive component is small, meaning the estimated diagonals on the $\mathbf{B}$ matrix are small.

Here is an example where we have both process and observation error but the covariates only affect the observation process:
\begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}\mathbf{x}_{t-1}  + \mathbf{w}_t, \text{ where } \mathbf{w}_t \sim \text{MVN}(0,\QQ)\\ 
\mathbf{y}_t = \mathbf{x}_{t-1} + \mathbf{D}\mathbf{d}_t  \mathbf{v}_t, \text{ where } \mathbf{v}_t \sim \text{MVN}(0,\mathbf{R}), 
\end{gathered}
(\#eq:msscov-xy-obs-only)
\end{equation}
$\mathbf{x}$ is the true algae abundances and $\mathbf{y}$ is the observation of the $\mathbf{x}$'s.

```{r msscov-covar-model-6 }
C <- c <- A <- U <- "zero"; Z <- "identity"
B <- "diagonal and unequal"
Q <- "equalvarcov"
D <- "unconstrained"
d <- covariates
R <- diag(0.16,2)
x0 <- "unequal"
tinitx <- 1
model.list <- list(B=B,U=U,Q=Q,Z=Z,A=A,R=R,D=D,d=d,C=C,c=c,x0=x0,tinitx=tinitx)
kem <- MARSS(dat, model=model.list)
```

## Including seasonal effects in MARSS models {#sec-msscov-season}

Time-series data are often collected at intervals with some implicit ``seasonality.'' For example, quarterly earnings for a business, monthly rainfall totals, or hourly air temperatures. In those cases, it is often helpful to extract any recurring seasonal patterns that might otherwise mask some of the other temporal dynamics we are interested in examining.

Here we show a few approaches for including seasonal effects using the Lake Washington plankton data, which were collected monthly.  The following examples will use all five phytoplankton species from Lake Washington.  First, let's set up the data.

```{r msscov-set-up-seasonal-dat}
years <- fulldat[,"Year"]>=1965 & fulldat[,"Year"]<1975
phytos <- c("Diatoms", "Greens", "Bluegreens",
           "Unicells", "Other.algae")
dat <- t(fulldat[years, phytos])

# z.score data because we changed the mean when we subsampled
the.mean <- apply(dat,1,mean,na.rm=TRUE)
the.sigma <- sqrt(apply(dat,1,var,na.rm=TRUE))
dat <- (dat-the.mean)*(1/the.sigma)
# number of time periods/samples
TT <- dim(dat)[2]
```


### Seasonal effects as fixed factors {#sec-msscov-season-factors}

One common approach for estimating seasonal effects is to treat each one as a fixed factor, such that the number of parameters equals the number of ``seasons'' (e.g., 24 hours per day, 4 quarters per year). The plankton data are collected monthly, so we will treat each month as a fixed factor.  To fit a model with fixed month effects, we create a $12 \times T$ covariate matrix $\mathbf{c}$ with one row for each month (Jan, Feb, ...) and one column for each time point. We put a 1 in the January row for each column corresponding to a January time point, a 1 in the February row for each column corresponding to a February time point, and so on. All other values of $\mathbf{c}$ equal 0. The following code will create such a $\mathbf{c}$ matrix.

```{r msscov-set-up-month-factors}
# number of "seasons" (e.g., 12 months per year)
period <- 12
# first "season" (e.g., Jan = 1, July = 7)
per.1st <- 1
# create factors for seasons
c.in <- diag(period)
for(i in 2:(ceiling(TT/period)))
  {c.in <- cbind(c.in,diag(period))}
# trim c.in to correct start & length
c.in <- c.in[,(1:TT)+(per.1st-1)]
# better row names
rownames(c.in) <- month.abb
```

Next we need to set up the form of the $\mathbf{C}$ matrix which defines any constraints we want to set on the month effects.  $\mathbf{C}$ is a $5 \times 12$ matrix. Five taxon and 12 month effects. 
If we wanted each taxon to have the same month effect, i.e. there is a common month effect across all taxon, then 
we have the same value in each $\mathbf{C}$ column\footnote{`month.abb` is a R constant that gives month abbreviations in text.}:

```{r msscov-C-constrained}
C <- matrix(month.abb,5,12,byrow=TRUE)
C
```

Notice, that $\mathbf{C}$ only has 12 values in it, the 12 common month effects.
However, for this example, we will let each taxon have a different month effect thus allowing different seasonality for each taxon.  For this model, we want each value in $\mathbf{C}$ to be unique:

```{r msscov-C-constrained2}
C <- "unconstrained"
```
Now $\mathbf{C}$ has 5 $\times$ 12 = 60 separate effects.

Then we set up the form for the rest of the model parameters.  We make the following assumptions:

```{r msscov-month-factor-marss-params}
# Each taxon has unique density-dependence
B <- "diagonal and unequal"
# Assume independent process errors
Q <- "diagonal and unequal"
# We have demeaned the data & are fitting a mean-reverting model
# by estimating a diagonal B, thus
U <- "zero"
# Each obs time series is associated with only one process
Z <- "identity" 
# The data are demeaned & fluctuate around a mean
A <- "zero" 
# We assume observation errors are independent, but they
# have similar variance due to similar collection methods
R <- "diagonal and equal"
# We are not including covariate effects in the obs equation
D <- "zero"
d <- "zero"
```

Now we can set up the model list for MARSS and fit the model (results are not shown since they are verbose with 60 different month effects).

```{r msscov-fit-month-factor-with-MARSS, results='hide'}
model.list <- list(B=B,U=U,Q=Q,Z=Z,A=A,R=R,C=C,c=c.in,D=D,d=d)
seas.mod.1 <- MARSS(dat,model=model.list,control=list(maxit=1500))

# Get the estimated seasonal effects
# rows are taxa, cols are seasonal effects
seas.1 <- coef(seas.mod.1,type="matrix")$C
rownames(seas.1) <- phytos
colnames(seas.1) <- month.abb
```

The top panel in Figure \@ref(fig:msscov-mon-effects) shows the estimated seasonal effects for this model.  Note that if we had set U="unequal", we would need to set one of the columns of $\mathbf{C}$ to zero because the model would be under-determined (infinite number of solutions). If we substracted the mean January abundance off each time series, we could set the January column in $\mathbf{C}$ to 0 and get rid of 5 estimated effects.

### Seasonal effects as a polynomial {#sec-msscov-season-polynomial}

The fixed factor approach required estimating 60 effects.  Another approach is to model the month effect as a 3rd-order (or higher) polynomial: $a+b\times m + c\times m^2 + d \times m^3$ where $m$ is the month number. This approach has less flexibility but requires only 20 estimated parameters (i.e., 4 regression parameters times 5 taxa).  To do so, we create a $4 \times T$ covariate matrix $\mathbf{c}$ with the rows corresponding to 1, $m$, $m^2$, and $m^3$, and the columns again corresponding to the time points.  Here is how to set up this matrix:

```{r msscov-poly-month-factor, results='hide'}
# number of "seasons" (e.g., 12 months per year)
period <- 12
# first "season" (e.g., Jan = 1, July = 7)
per.1st <- 1
# order of polynomial
poly.order <- 3
# create polynomials of months
month.cov <- matrix(1,1,period)
for(i in 1:poly.order) {month.cov = rbind(month.cov,(1:12)^i)}
# our c matrix is month.cov replicated once for each year
c.m.poly <- matrix(month.cov, poly.order+1, TT+period, byrow=FALSE)
# trim c.in to correct start & length
c.m.poly <- c.m.poly[,(1:TT)+(per.1st-1)]

# Everything else remains the same as in the previous example
model.list <- list(B=B,U=U,Q=Q,Z=Z,A=A,R=R,C=C,c=c.m.poly,D=D,d=d)
seas.mod.2 <- MARSS(dat, model=model.list, control=list(maxit=1500))
```
The effect of month $m$ for taxon $i$ is $a_i + b_i \times m + c_i \times m^2 + d_i \times m^3$, where $a_i$, $b_i$, $c_i$ and $d_i$ are in the $i$-th row of $\mathbf{C}$.  We can now calculate the matrix of seasonal effects as follows, where each row is a taxon and each column is a month:
```{r msscov-seasonal-effect-poly}
C.2 = coef(seas.mod.2,type="matrix")$C
seas.2 = C.2 %*% month.cov
rownames(seas.2) <- phytos
colnames(seas.2) <- month.abb
```
The middle panel in Figure \@ref(fig:msscov-mon-effects) shows the estimated seasonal effects for this polynomial model.

Note: Setting the covariates up like this means that our covariates are collinear since $m$, $m^2$ and $m^3$ are correlated, obviously.  A better approach is to use the `poly()` function to create an orthogonal polynomial covariate matrix `c.m.poly.o`:
```{r msscov-poly}
month.cov.o <- cbind(1, poly(1:period, poly.order))
c.m.poly.o <- matrix(t(month.cov.o), poly.order+1, TT+period, byrow=FALSE)
c.m.poly.o <- c.m.poly.o[,(1:TT)+(per.1st-1)]
```


### Seasonal effects as a Fourier series {#sec-msscov-season-fourier}

The factor approach required estimating 60 effects, and the 3rd order polynomial model was an improvement at only 20 parameters. A third option is to use a discrete Fourier series, which is combination of sine and cosine waves; it would require only 10 parameters. Specifically, the effect of month $m$ on taxon $i$ is $a_i \times \cos(2 \pi m/p) + b_i \times \sin(2 \pi m/p)$, where $p$ is the period (e.g., 12 months, 4 quarters), and $a_i$ and $b_i$ are contained in the $i$-th row of $\mathbf{C}$.

We begin by defining the $2 \times T$ seasonal covariate matrix $\mathbf{c}$ as a combination of 1 cosine and 1 sine wave:
```{r msscov-seasonal-fourier}
cos.t <- cos(2 * pi * seq(TT) / period)
sin.t <- sin(2 * pi * seq(TT) / period)
c.Four <- rbind(cos.t,sin.t)
```

Everything else remains the same and we can fit this model as follows:
```{r msscov-seasonal-fourier-fit, results='hide'}
model.list <- list(B=B,U=U,Q=Q,Z=Z,A=A,R=R,C=C,c=c.Four,D=D,d=d)
seas.mod.3 <- MARSS(dat, model=model.list, control=list(maxit=1500))
```

We make our seasonal effect matrix as follows:
```{r msscov-seasonal-effects-fourier}
C.3 <- coef(seas.mod.3, type="matrix")$C
# The time series of net seasonal effects
seas.3 <- C.3 %*% c.Four[,1:period]
rownames(seas.3) <- phytos
colnames(seas.3) <- month.abb
```
The bottom panel in Figure \@ref(fig:msscov-mon-effects) shows the estimated seasonal effects for this seasonal-effects model based on a discrete Fourier series.

(ref:msscov-mon-effects) Estimated monthly effects for the three approaches to estimating seasonal effects. Top panel: each month modelled as a separate fixed effect for each taxon (60 parameters); Middle panel: monthly effects modelled as a 3rd order polynomial (20 parameters); Bottom panel: monthly effects modelled as a discrete Fourier series (10 parameters).

```{r msscov-mon-effects, fig=TRUE, echo=FALSE, fig.cap='(ref:msscov-mon-effects)', warning=FALSE}
par(mfrow=c(3,1), mar=c(2,4,2,2)) 
matplot(t(seas.1), type="l", bty="n", xaxt="n", ylab="Fixed monthly", col=1:5)
axis(1, labels=month.abb, at=1:12, las=2, cex.axis=0.75)
legend("topright", lty=1:5, legend=phytos, cex=0.6, col=1:5)

matplot(t(seas.2),type="l",bty="n",xaxt="n", ylab="Cubic", col=1:5)
axis(1,labels=month.abb, at=1:12,las=2,cex.axis=0.75)
legend("topright", lty=1:5, legend=phytos, cex=0.6, col=1:5)

matplot(t(seas.3),type="l",bty="n",xaxt="n",ylab="Fourier", col=1:5)
axis(1,labels=month.abb, at=1:12,las=2,cex.axis=0.75)
legend("topright", lty=1:5, legend=phytos, cex=0.6, col=1:5)
```

Rather than rely on our eyes to judge model fits, we should formally assess which of the 3 approaches offers the most parsimonious fit to the data. Here is a table of AICc values for the 3 models:
```{r msscov-show-aics}
data.frame(Model=c("Fixed", "Cubic", "Fourier"),
           AICc=round(c(seas.mod.1$AICc,
                        seas.mod.2$AICc,
                        seas.mod.3$AICc),1))
```
The model selection results indicate that the model with monthly seasonal effects estimated via the discrete Fourier sequence is the best of the 3 models. Its AICc value is much lower than either the polynomial or fixed-effects models.

## Model diagnostics {#sec-msscov-model-diagnostics}

We will examine some basic model diagnostics for these three approaches by looking at plots of the model residuals and their autocorrelation functions (ACFs) for all five taxa using the following code:

```{r msscov-diagnostic-code, eval=FALSE}
for(i in 1:3) {
  dev.new()
  modn <- paste("seas.mod",i,sep=".")
  for(j in 1:5) {
    plot.ts(residuals(get(modn))$model.residuals[j,], 
      ylab="Residual", main=phytos[j])
    abline(h=0, lty="dashed")
    acf(residuals(get(modn))$model.residuals[j,])
    }
  }
```

(ref:msscov-diagnostic-fig) Residuals for model with season modelled as a discrete Fourier series.

```{r msscov-diagnostic-fig, fig=TRUE, echo=FALSE, fig.cap='(ref:msscov-diagnostic-fig)', warning=FALSE}
  i <- 3; #Fourier
  j <- 1; #First state
  par(mfrow=c(2,1))
  modn <- paste("seas.mod",i,sep=".")
  plot.ts(residuals(get(modn))$model.residuals[j,], 
      ylab="Residual", main=phytos[j])
  abline(h=0, lty="dashed")
  acf(residuals(get(modn))$model.residuals[j,])
```

## Homework data and discussion {#sec-msscov-hw-discussion}

For these problems, use the following code to load in 1980-1994 phytoplankton data, covariates, and z-score all the data.  Run the code below and use `dat` and `covars` directly in your code.
```{r msscov-problems-dat, results="hide"}
library(MARSS)
spp <- c("Cryptomonas", "Diatoms", "Greens",
            "Unicells", "Other.algae","Daphnia")
yrs <- lakeWAplanktonTrans[,"Year"]%in%1980:1994
dat <- t(lakeWAplanktonTrans[yrs,spp])
#z-score the data
avg <- apply(dat, 1, mean, na.rm=TRUE)
sd <- sqrt(apply(dat, 1, var, na.rm=TRUE))
dat <- (dat-avg)/sd
rownames(dat)=spp
#always check that the mean and variance are 1 after z-scoring
apply(dat,1,mean, na.rm=TRUE) #this should be 0
apply(dat,1,var, na.rm=TRUE) #this should be 1
```

For the covariates, you'll use temperature and TP.

```{r msscov-problems-covar, results="hide"}
covars <- rbind(Temp=lakeWAplanktonTrans[yrs,"Temp"],
               TP=lakeWAplanktonTrans[yrs,"TP"])
avg <- apply(covars, 1, mean)
sd <- sqrt(apply(covars, 1, var, na.rm=TRUE))
covars <- (covars-avg)/sd
rownames(covars) <- c("Temp","TP")
#always check that the mean and variance are 1 after z-scoring
apply(covars,1,mean, na.rm=TRUE) #this should be 0
apply(covars,1,var, na.rm=TRUE) #this should be 1
```

Here are some guidelines to help you answer the questions:

* Use a MARSS model that allows for both observation and process error.
* Assume that the observation errors are independent and identically distributed with known variance of 0.10. 
* Assume that the process errors are independent from one another, but the variances differ by taxon.
* Assume that each group is an observation of its own process. This means `Z="identity"`.
* Use `B="diagonal and unequal"`.  This implies that each of the taxa are operating under varying degrees of density-dependence, and they are not allowed to interact.
* All the data have been de-meaned and $\mathbf{Z}$ is identity, therefore use `U="zero"` and `A="zero"`.  Make sure to check that the means of the data are 0 and the variance is 1.
* Use `tinitx=1`. It makes $\mathbf{B}$ estimation more stable. It goes in your model list.
* Include a plot of residuals versus time and acf of residuals for each question. You only need to show these for the top (best) model if the question involves comparing multiple models.
* Use AICc to compare models.
* Some of the models may not converge, however use for the purpose of the homework, use the unconverged models.  Thus use the output from `MARSS()` without any additional arguments.  If you want, you can try using `control=list(maxit=1000)` to increase the number of iterations.  Or you can try `method="BFGS"` in your `MARSS()` call. This will use the BFGS optimization method, however it may throw an error for these data.

\clearpage

## Problems {#sec-msscov-problems}

Read Section \@ref(sec-msscov-hw-discussion) for the data and tips on answering the questions and setting up your models.  Note the questions asking about the effects on *growth rate* are asking about the _C_ matrix in 
$$\mathbf{x}_t=\mathbf{B}\mathbf{x}_{t-1}+\mathbf{C}\mathbf{c}_t+\mathbf{w}_t$$
The $\mathbf{C}\mathbf{c}_t+\mathbf{w}_t$ are the process errors and represent the growth rates (growth above or below what you would expect given $\mathbf{x}_{t-1}$).  Use your raw data in the MARSS model.  You do not need to difference the data to get at the growth rates since the process model is modeling that.

1. How does month affect the mean phytoplankton population growth rates? Show a plot of the estimated mean growth rate versus month for each taxon using three approaches to estimate the month effect (factor, polynomial, Fourier series). Estimate seasonal effects without any covariate (Temp, TP) effects.

2. It is likely that both temperature and total phosphorus (TP) affect phytoplankton population growth rates. Using MARSS models, estimate the effect of Temp and TP on growth rates of each taxon. 
Leave out the seasonal covariates from question 1, i.e. only use Temp and TP as covariates. Make a plot of the point estimates of the Temp and TP effects with the 95% CIs added to the plot. `tidy()` is an easy way to get the parameters CIs.

3. Estimate the Temp and TP effects using `B="unconstrained"`.  
    a. Compare the $\mathbf{B}$ matrix for the fit from question 2 and from question 3.  Describe the species interactions modeled by the $\mathbf{B}$ matrix when `B="unconstrained"`.  How is it different than the $\mathbf{B}$ matrix from question 2? Note, you can retrieve the matrix using `coef(fit, type="matrix")$B`.
    b. Do the Temp and TP effects change when you use `B="unconstrained"`? Make sure to look at the CIs also.

4. Using MARSS models, evaluate which (Temp or TP) is the more important driver or if both are important. Again, leave out the seasonal covariates from question 1, i.e. only use Temp and TP as covariates. Compare two approaches: comparison of effect sizes in a model with both Temp and TP and model selection using a set of models.

5. Evaluate whether the effect of temperature (Temp) on the taxa manifests itself via their underlying physiology (by affecting growth rates and thus abundance) or because physical changes in the water stratification makes them easier/harder to sample in some months. Leave out the seasonal covariates from question 1, i.e. only use Temp and TP as the covariates. For TP, assume it always affects growth rates, never the observation errors.

6. Is there support for temperature or TP  affecting all functional groups' growth rates the same, or are the effects on one taxon different from another? Make sure to test all possibilities: the Temp and TP effects are the same for all taxa, and one covariate effect is the same across taxa  while the other's effects are unique across taxa.

7. Compare your results for question 2 using linear regression, by using the `lm()` function. You'll need to look at the response of each taxon separately, i.e. one response variable.  You can have a multivariate response variable with `lm()` but the functions will be doing 6 independent linear regressions. In your `lm()` model, use only Temp and TP (and intercept) as covariates.  Compare the estimated effects to those from question 2.  How are they different? How is this model different from the model you fit in question 2?

<!--

8. Using Greens only, fit a linear regression with autocorrelated residuals using the `Arima()` function with the `xreg` argument. Use Temp and TP as your covariates.  Compare your results to  the results from questions 7. How is this model different from the models you fit in questions 7?

-->

8. Temp and TP are negatively correlated (cor = `r format(cor(covars[1,],covars[2,]), digits=2)`). A common threshold for collinearity in regression models is 0.7. Temp and TP fall below that but are close. One approach to collinearity is sequential regression [@Dormannetal2013]. The first (most influential) covariate is included 'as is' and the second covariate appears as the residuals of a regression of the second against the first. The covariates are now orthogonal however the second covariate is conditioned on the first. If we see an effect of the residuals covariate, it is the effect of TP additional  to  the  contribution  it  already  made  through  its  relationship  with  temperature. Rerun question 2 using sequential regression (see code below).  

Make your Temp and TP covariates orthogonal using sequential regression. Do your conclusions about the effects of Temperature and TP change?

Below is code to construct your orthogonal covariates for sequential regression.
```{r msscov-problems-orthocov}
fit <- lm(covars[1,]~covars[2,])
seqcovs <- rbind(covars[1,], residuals(fit))
avg <- apply(seqcovs, 1, mean)
sd <- sqrt(apply(seqcovs, 1, var, na.rm=TRUE))
seqcovs <- (seqcovs-avg)/sd
rownames(seqcovs) <- c("Temp","TPresids")
```

9. Compare the AICc's of the 3 seasonal models from question 1 and the 4 Temp/TP models from question 5. What does this tell you about the Temp and TP only models?

10. We cannot just fit a model with season and Temp plus TP since Temp and TP are highly seasonal.  That will cause problems if we have something that explain season (a polynomial) and a covariate that has seasonality.  Instead, use sequential regression to fit a model with seasonality, Temp and TP. Use a 3rd order polynomial with the `poly()` function to create orthogonal season covariates and then use sequential regression (code in problem 8) to create Temp and TP covariates that are orthogonal to your season covariates.  Fit the model and compare a model with only season to a model with season and Temp plus TP.

11. Another approach to looking at effects of covariates which have season cycles is to examine if the seasonal anomalies of the independent variable can be explained by the seasonal anomalies of the dependent variables.  In other words, can an unusually high February abundance (higher than expected) be explained by an unusually high or low February temperature? In this approach, you remove season so you do not need to model it (with factor, polynomial, etc). The `stl()` function can be used to decompose a time series using LOESS. We'll use `stl()` since it can handle missing values. 

    a. Decompose the Diatom time series using `stl()` and plot. Use `na.action=zoo::na.approx` to deal with the NAs. Use `s.window="periodic"`. Other than that you can use the defaults.
```{r, eval=FALSE}
i <- "Diatoms"
dati <- ts(dat[i,], frequency=12)
a <- stl(dati, "periodic", na.action=zoo::na.approx)
```

    b. Create dependent variables and covariates that are anomolies by modifying the following code. For the anomaly, you will use the remainder plus the trend. You will need to adapt this code to create the anomalies for Temp and TP and for `dat` (your data).
```{r}
i <- "Diatoms"
a <- stl(ts(dat[i,],frequency=12), "periodic", na.action=zoo::na.approx)
anom <- a$time.series[,"remainder"]+a$time.series[,"trend"]
```

    c. Notice that you have simply removed the seasonal cycle from the data. Using the seasonal anomalies (from part b), estimate the effect of Temp and TP on each taxon's growth rate. You will use the same model as in question 2, but use the seasonal anomalies as data and covariates.



```{r}
anoms <- matrix(NA,dim(dat)[1]+dim(covars)[1],dim(dat)[2])
rownames(anoms) <- c(rownames(dat), rownames(covars))
for(i in 1:dim(dat)[1]){
   a <- stl(ts(dat[i,],frequency=12), "periodic", na.action=zoo::na.approx)
   anoms[i,] <- a$time.series[,"remainder"]+a$time.series[,"trend"]
}
for(i in 1:dim(covars)[1]){
   a <- stl(ts(covars[i,],frequency=12), "periodic", na.action=zoo::na.approx)
   anoms[i+dim(dat)[1],] <- a$time.series[,"remainder"]+a$time.series[,"trend"]
}
```


