## Problems {#sec-week7-problems}

For the second part of the homework, we'll use data from the Pacific Decadal Oscillation (PDO) to ask questions about identifying regimes. This dataset can be accessed via the `rpdo` package. First, let's grab the data. 

```{r read-data}
library(dplyr)
install.packages("rsoi")
pdo <- rsoi::download_pdo()
pdo$water_year = pdo$Year
pdo$water_year[which(pdo$Month%in%c("Oct","Nov","Dec"))] = pdo$water_year[which(pdo$Month%in%c("Oct","Nov","Dec"))] + 1
pdo = dplyr::group_by(pdo, water_year) %>%
  dplyr::summarize(winter_pdo = mean(PDO[which(Month %in% c("Oct","Nov","Dec","Jan","Feb"))])) %>% 
  dplyr::select(winter_pdo,water_year) %>% 
  dplyr::rename(year=water_year)
```

1. Identifying regimes using Hidden Markov Models (HMMs)

    a. Start by fitting a 2-state HMM to the annual indices of winter PDO. You're welcome to use any package / method you like ('depmixS4' would be a good choice if you're unsure). Assume Gaussian errors. 
    
    b. Try to fit the model 10-20 times. Does the likelihood seem reasonably stable?
    
    c. Change the model to a 3-state model. Using AIC as a model selection metric, does the 3-state model perform better (lower AIC) compared to the 2-state model? What about a 1-state model?
    
    d. What is the transition matrix for the best model? What are the persistence probabilities (e.g. probabilities of being in the same state)?
    
    e. Plot the probability of being in the various states from your best model (e.g. probability of being in state 1 over time)
    
    f. Plot the time series of predicted values form the model
    
    f. If you include time varying parameters (e.g. year) in the means of each state, or state transition probabilities, does the model do any better?
    
2. Bayesian MARSS modelling 

```{r}
data(neon_barc, package = "atsalibrary")
data <- neon_barc
data$indx <- seq(1, nrow(data))
n_forecast <- 7
n_lag_o2 <- 1
n_lag_temp <- 1
last_obs <- nrow(data)

create_stan_data <- function(data, last_obs, n_forecast, n_lag_o2, 
    n_lag_temp) {
    # create test data
    o2_test <- dplyr::filter(data, indx %in% seq(last_obs + 1, 
        (last_obs + n_forecast)))
    temp_test <- dplyr::filter(data, indx %in% seq(last_obs + 
        1, (last_obs + n_forecast)))
    
    o2_train <- dplyr::filter(data, indx <= last_obs, !is.na(oxygen))
    o2_x <- o2_train$indx
    o2_y <- o2_train$oxygen
    o2_sd <- o2_train$oxygen_sd
    n_o2 <- nrow(o2_train)
    
    temp_train <- dplyr::filter(data, indx <= last_obs, !is.na(temperature))
    temp_x <- temp_train$indx
    temp_y <- temp_train$temperature
    temp_sd <- temp_train$temperature_sd
    n_temp <- nrow(temp_train)
    
    stan_data <- list(n = last_obs, n_lag_o2 = n_lag_o2, n_lag_temp = n_lag_temp, 
        n_forecast = n_forecast, n_o2 = n_o2, o2_x = o2_x, o2_y = o2_y, 
        o2_sd = o2_sd, n_temp = n_temp, temp_x = temp_x, temp_y = temp_y, 
        temp_sd = temp_sd)
    
    return(list(o2_train = o2_train, temp_train = temp_train, 
        stan_data = stan_data, o2_test = o2_test, temp_test = temp_test))
}
m <- stan_model(file = "model_02.stan")
```





